---
title: "Geopackage"
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

```{r}
#| label: setup

# library(gdalvector)
require(gdalraster)
```

## Overview

::: callout-note
### Useful Links

- [GPKG - GDAL Vector Driver](https://gdal.org/en/stable/drivers/vector/gpkg.html)
- [GeoPackage Standard – Open SQLite Format for Geospatial Data](https://www.ogc.org/standards/geopackage/)
- [OGC GeoPackage Encoding Standard](https://www.geopackage.org/spec140/index.html)
- [OGC GeoPackage Format standard](http://www.geopackage.org/spec/)
- [Getting Started With GeoPackage](http://www.geopackage.org/guidance/getting-started.html)
- [SQLite - GDAL Vector Driver](https://gdal.org/en/stable/drivers/vector/sqlite.html)
- [SQLite Home Page](https://www.sqlite.org/index.html)
- [Pragma Statements supported by SQLite](https://www.sqlite.org/pragma.html)
- [SpatiaLite](https://www.gaia-gis.it/fossil/libspatialite/index)
- [Geo tips & tricks: Dealing with huge vector GeoPackage databases in GDAL/OGR](https://erouault.blogspot.com/2017/03/dealing-with-huge-vector-geopackage.html?m=1)
:::

The [GeoPackage](https://www.ogc.org/standards/geopackage/) (`GPKG`) format format is an open, non-proprietary, 
platform-independent and standards-based data format for geographic information system implemented as a 
[SQLite Database](https://sqlite.org) using the [`.gpkg`](https://www.geopackage.org/spec120/#r3) file extension.

## Identity

A GeoPackage is fundamentally a SQLite database and can be opened as such with any SQLite client. 

The SQLite container itself is identified by the first 16 bytes of the file, which must match the null-terminated ASCII
string [`"SQLite format 3\0"`](https://www.geopackage.org/spec120/#r1). 

Everything beyond that is governed by the GeoPackage standard, which layers two additional identity fields on top of
the SQLite header: the `application_id` and `user_version` PRAGMAs at byte offsets 68 and 60, respectively.

For verification of the identity of a particular gpkg file, the following criteria should be checked:

- [ ] The file must have a `.gpkg` extension (case-insensitive).

- [ ] The file must be a valid SQLite database, which can be verified by checking the first 16 bytes,
  which must match the null-terminated ASCII string `"SQLite format 3\0"`:
  
"SQLite format 3\0" in bytes:

```plaintext
53 51 4c 69 74 65 20 66 6f 72 6d 61 74 20 33 00
```

- [ ] The `PRAGMA application_id;` must contain a recognized GeoPackage application identifier:

|  ID String | 32-bit Integer |  Hexadecimal |
| :--------: | :------------: | :----------: |
| `"GP10"`   | `1196437808`   | `0x47503130` |
| `"GP11"`   | `1196437809`   | `0x47503131` |
| `"GPKG"`   | `1196444487`   | `0x47504B47` |

- [ ] The `PRAGMA user_version;` must contain a recognized GeoPackage version identifier, encoded as 
  `<major> x 10000 + <minor> x 100 + <bugfix/patch>` (integer with major version, two-digit minor version, 
  and two-digit "bug-fix" version), i.e. `10200` for "GeoPackage 1.2":

| Version (String) | Version Integer |  Hexadecimal  |
| :--------------: | :------------: | :---------:    |
| `"GPKG 1.2"`     | `10200`        | `0x000027D8`   |
| `"GPKG 1.3"`     | `10300`        | `0x0000283C`   |
| `"GPKG 1.4"`     | `10400`        | `0x000028A0`   |

::: callout-note

**Notes**:

- All hexadecimal representations are using big-endian byte order, consistent with the SQLite file format specification.

- Both `application_id` and `user_version` are stored as 32-bit integers at fixed offsets in the SQLite file header
  (offsets 68 and 60 respectively), and their hexadecimal values reflect this storage format.

- GeoPackage versions `1.0`, `1.0.1`, and `1.1` predate the `user_version` mechanism and use `application_id` alone for 
  version identification (`"GP10"` and `"GP11"` respectively). The `PRAGMA user_version;` check applies only to 
  GeoPackage 1.2 and later, all of which use `application_id = "GPKG"` (`0x47504B47`).
  
- The `application_id` can be set programmatically via `PRAGMA application_id = 1196444487;` (SQLite ≥ `3.7.17`), or by
  writing the byte sequence `47 50 4B 47` directly at offset 68 in the file header for older SQLite versions.
:::

The following code demonstrates reading and verifying these identity fields directly:

```{r}
#| label: identity-sqlite-header

# set the path for an example gpkg to use:
gpkg_path <- pkg_example_data("ssurgo/ssurgo.gpkg")

# read the 16-byte SQLite magic header:
read_magic_header(gpkg_path, n = 16L)
```

The following enumerated constants are used internally in `gdalvector` to perform identity checks on the `application_id`
and `user_version` PRAGMAs for a provided `gpkg`:

```{r}
#| label: identity-constants

# application_ids are always checked and should always equal the "GPKG" application_id below, unless the application_id
# is used as the versioning mechanism, i.e. for GeoPackage 1.0, GeoPackage 1.0.1 and GeoPackage 1.1, 
# the application_id is used to determine the version, and the user_version is not used:
GPKG_APPLICATION_IDS <- c("GPKG" = char_to_int("GPKG"), "GP10" = char_to_int("GP10"), "GP11" = char_to_int("GP11"))

# only need to check user_version for GPKG 1.2 and later, for earlier versions the application_id is used to determine the version
# will almost always be 1.2:
GPKG_USER_VERSIONS <- c("GPKG 1.2" = 10200L, "GPKG 1.3" = 10300L, "GPKG 1.4" = 10400L)
```

These can be verified via a direct SQLite connection:

```{r}
#| label: identity-sqlite-pragmas

gpkg_pragma(gpkg_path, pragma = "application_id")
gpkg_pragma(gpkg_path, pragma = "user_version")
```

or through GDAL using `gdal vector sql` with the `SQLITE` dialect:

```{r}
#| label: gpkg-pragmas-gdal

# get the application_id pragma through GDAL:
gdal_vector_sql(gpkg_path, layer = gdal_vector_layer(gpkg_path), sql = "PRAGMA application_id;", dialect = "SQLITE")

# get the user_version pragma through GDAL:
gdal_vector_sql(gpkg_path, layer = gdal_vector_layer(gpkg_path), sql = "PRAGMA user_version;", dialect = "SQLITE")
```






  
